DeepPicar:​ ​A​ ​Low-cost​ ​Deep​ ​Neural​ ​Network-based​ ​Autonomous​ ​Car

### 2017-10-11 paper/appendix.tex
intro revision
Minor changes to paper

\subsection{DNN Training and Testing}
We have trained and tested the deep neural network with several
different track conditions, different combinations of input
data, and different hyper parameters. In the following, we describe
details on two of the training methods that performed reasonably well.

@@

We continue to investigate ways to achieve better prediction accuracy
in training the network, as well as improving the performance of the RC
car platform, especially on precise steering angle control.

>>>>

\subsection{DNN Training and Testing}
We have trained and tested the deep neural network with several
different track conditions, different combinations of input
data, and different hyper parameters. In the following paragraphs, we 
describe details on two of the training methods that performed 
reasonably well.

@@

We continue to investigate ways to achieve better prediction accuracy
in training the network, as well as improving the performance of the RC
car platform, especially related to precise steering angle control.

### 2017-10-11 paper/comparison.tex
Minor changes to paper

Figure~\ref{fig:sys_core} shows the average control loop completion
timing of the four system configurations we tested as a function of
the number of CPU cores used.
First, both the Intel UP and Tegra TX2 exhibit superior performance when
compared with the Raspberry Pi 3. 
When all four CPU cores are used, the Intel UP is 2.53X faster than
the Pi 3, while TX (CPU) and TX (GPU) are 3.6X and 7.6X times faster,
respectively, than the Pi 3. 
As a result, they are all able to satisfy the 50 ms 
WCET by a clear margin
%%  (in the worst case, the UP Board would 
%% still meet it by $\sim$20 ms) --> where's the data showing this?
, and, in case of TX2, 50 Hz or even 100 Hz real-time control is
feasible with the help of its GPU. Another observation is that TX2
(GPU) does not change much as most of the neural network computation
is done at the GPU.

@@

Finally, we compare the effect of co-scheduling memory bandwidth
intensive synthetic benchmarks on the DNN control loop timing. 
Figure~\ref{fig:sys_bench} shows the results.
As discussed in 
Section~\ref{sec:eval-memhog}, we observe dramatic execution time
increase, up to 9.4X, in Raspberry Pi 3 as we increase the number of
co-scheduled tasks. We also observe increased control loop execution
timing in Intel Up and TX2, but the degree of the increase is not as
dramatic as the Pi3. Compared to their respective solo timings (i.e.,
the model runs on a single core in isolation), Intel UP suffers up to
2.3X execution time increase; TX2 (CPU) and TX2 (GPU) suffer up to
2.2X and 2.5X increase, respectively. This is somewhat suprising
because Raspberry Pi 3's cores are in-order architecture based while
the cores in Intel Up and NVIDIA TX2 are out-of-order architecture
based and that the memory intensive tasks on out-of-order cores can
generate more memory traffic. We suspect that this is because the
memory subsystems in the Intel UP and TX2 platforms provide higher
bandwidth and fairness than the memory subsystem of the Pi 3.

Another intersting observation is that TX2 (GPU) also suffers
considerable execution time increase (2.5X) despite the fact that the
co-scheduled synthetic tasks do not utilize the GPU. In other words,
the DNN model has dedicated access to the GPU. This is, however, a
known characteristic of integrated CPU-GPU architectue based
platforms that both CPU and GPU share the same memory
subsystem~\cite{Ali2017}. As a result, TX2 (GPU) fails to meet the
10ms deadline for 100 Hz control that would have been feasible if
there was no contention between the CPU cores and the GPU.

>>>>

Figure~\ref{fig:sys_core} shows the average control loop completion
timing of the four system configurations we tested as a function of
the number of CPU cores used.
First, both the Intel UP and Tegra TX2 exhibit superior performance when
compared with the Raspberry Pi 3. 
When all four CPU cores are used, the Intel UP is 2.53X faster than
the Pi 3, while TX (CPU) and TX (GPU) are 3.6X and 7.6X times faster,
respectively, than the Pi 3. 
As a result, they are all able to satisfy the 50 ms 
WCET by a clear margin,
%%  (in the worst case, the UP Board would 
%% still meet it by $\sim$20 ms) --> where's the data showing this?
and, in case of TX2, 50 Hz or even 100 Hz real-time control is
feasible with the help of its GPU. Another observation is that TX2
(GPU) does not change much, as most of the neural network computation
is done at the GPU.

@@

Finally, we compare the effect of co-scheduling memory bandwidth
intensive synthetic benchmarks on the DNN control loop timing. 
Figure~\ref{fig:sys_bench} shows the results.
As discussed in 
Section~\ref{sec:eval-memhog}, we observed dramatic execution time
increases, up to 9.4X, in Raspberry Pi 3 as we increased the number of
co-scheduled tasks. We also observe increased control loop execution
timing in Intel Up and TX2, but the degree of the increase is not as
dramatic as the Pi3. Compared to their respective solo timings (i.e.,
the model runs on a single core in isolation), Intel UP suffers up to
2.3X execution time increase; TX2 (CPU) and TX2 (GPU) suffer up to
2.2X and 2.5X increase, respectively. This is somewhat suprising
because Raspberry Pi 3's cores are in-order architecture based while
the cores in Intel Up and NVIDIA TX2 are out-of-order architecture
based, and that the memory intensive tasks on out-of-order cores can
generate more memory traffic. We suspect that this is because the
memory subsystems in the Intel UP and TX2 platforms provide higher
bandwidth and fairness than the memory subsystem of the Pi 3.

Another intersting observation is that TX2 (GPU) also suffers
considerable execution time increase (2.5X) despite the fact that the
co-scheduled synthetic tasks do not utilize the GPU. In other words,
the DNN model has dedicated access to the GPU. This is, however, a
known characteristic of integrated CPU-GPU architectue based
platforms where both CPU and GPU share the same memory
subsystem~\cite{Ali2017}. As a result, TX2 (GPU) fails to meet the
10ms deadline for 100 Hz control that would have been feasible if
there was no contention between the CPU cores and the GPU.

### 2017-10-11 paper/evaluation.tex
Minor changes to paper

For real-time control of a car (or any robots), the control loop
frequency must be sufficiently high so that the car can quickly
react to the changing environment and its internal states. In general,
control performance improves when the frequency is higher, though
computation time and the type of particular physical system are
factors in determining a proper control loop latency. While a standard
control system may be comprized of multiple control loops with
differning control frequencies---e.g., a inner control loop for lower-level
PD control, a outer loop for motion planning, etc.---DeepPicar's
control loop is a single layer, as shown earlier in
Figure~\ref{fig:controlloop}, as a single deep neural network
replaces the traditional multi-layer control pipline. (Recall
Figure~\ref{fig:end-to-end-control} on the differences between the
standard robotics control vs. end-to-end deep learning approach).
This means that the DNN interference operation must be completed
within the inner-most control loop update frequency. To understand
achievable control-loop update frequencies, we experimentally measured
the execution times of DeepPicar's DNN inference operations.

@@

First, as expected, we find that the inference operation
dominates the control loop execution time, accounting for about 85\% of
the execution time.
Second, more importantly, we also find that the measured average
execution time of a single control loop is 42.67 ms, or 23.4 Hz and
the 99 percentail time is 50.70ms.
This means that the DeepPicar can operate
at about 20 Hz control frequency in real-time using only the on-board
Raspberry Pi 3 computing platform, without needing any remote computing
resources. We consider these results respectable given the complexity
of the deep neural network, and the fact that the inference operation
performed by Tensorflow only utilizes the CPU cores of the
Raspberry Pi 3 (its GPU is not supported by Tensorflow.)

@@

In this experiment, we investigate the scalability of performing
inference operations of DeepPicar's neural network with respect to the
number of cores. As noted earlier, the Raspberry Pi 3 platform has
four Cortex-A53 cores and Tensorflow 
provides a programmable mechanism to to adjust how many cores are to be
used by the library. Leveraging this feature, we repeat the
same experiment described in the previous subsection with varying
numbers of CPU cores---from one to four.

@@

Figure~\ref{fig:perf-vs-corecnt} shows the average execution time of
the control loop as we vary the number of cores used by
Tensorflow. As expected, as we assign more cores, the average execution
time decreases---from 61.96 ms on a single core to 42.67 ms on four
cores (a 30\% improvement). However, the improvement is far from ideal
linear scaling. In particular, from 2 cores to 3 cores, the
improvement is mere 2.38ms (or 4\%). In short, we find that the
scalability of DeepPicar's deep neural network is not ideal on the
platform. We do not know whether it is due to the limitations of
Tensorflow's multicore implementation or it's the model's inherent
characteristics.

The poor scalability opens up the possibility of consolidating
multiple different tasks or different neural network models rather
than allocating all cores for a single neural network model. For
example, it is conceivable to use four cameras and four different
neural network models, each of which is trained separately and
executed on a single dedicated core. Assuming we use the same network
architecture for all models, then one might expect to achieve up to
15 Hz using one core (given 1 core can deliver 62ms average
execution time). In the next experiment, we investigate the
feasibility of such a scenario.

@@

Figure~\ref{fig:perf-vs-modelcnt} shows the results. In the figure, the
X-axis shows the system configuration: \#of DNN models x \#of CPU
cores/DNN. For example, '4x1' means running four DNN models each of
which is assigned to run on one CPU core, whereas '2x2' means running
two DNN models, each of which is assigned to run on two CPU
cores. The Y-axis shows the average inference timing.
Two bars in the left show the impact of co-scheduling four DNN
models. Compared to executing a single DNN model on one CPU core
(1x1), when four DNN models are co-scheduled (4x1), each model
suffers an average inference time increase of approximately 15 ms,
$\sim$24\%. On the other hand, when two DNN models, each using two CPU
cores, are co-scheduled, the average inference timing is increased by
about 7 ms, or 10\%, compared to the baseline of running one model
using two CPU cores.

@@

These increases in inference times in the co-scheduled scenarios are
expected and likely caused by contention in the shared hardware
resources such as the shared L2 cache and/or the DRAM controller.
To further analyze the source of contention, we use hardwawre
performance counters of the processor. Specifically, we measure L2
miss rates of the DNN models first in isolation and then after
co-scheduling other models. If the shared L2 cache is the primary
source of inteference, then the measured L2 miss rates would
increase. Figure~\ref{fig:l2missrate-vs-modelcnt} shows the results.
As can be see in the figure, L2 miss rates are largely unchanged
regardless whether multiple models are co-scheduled or not. This
suggests that the shared L2 cache is not the bottleneck that caused
execution time increases. In other words, DNN models don't appear to be
sensitive to the shared L2 cache space. % L2 partitioning is probably
                                % not going to be useful.
Instead, we hypothesize that it is likely caused by the memory
controller---the only other major shared hardware source---where
memory requests from different CPU cores contend, which would result
in increased memory access latency. While some Intel processors
provide oncore hardware counters that can measure average memory
access latency~\cite{ye2016maracas}, we were not able to identify
whether such hardware counters exist in the BCM2837 processor of
Raspberry Pi 3 due to the lack of documentation. Instead, in the next
experiment, we use memory intensive synthetic benchmarks to test the
hypothesis.

@@

Figure~\ref{fig:} shows the normalized execution time and L2 miss-rate
of the DNN model running on the Core 0 as a function of the number of
co-scheduled memory intensive synthetic benchmarks. First, as we
increase the number of co-runners, the DNN model's execution times are
increased---by up to 9.6X---even though the DNN model is running on a
dedicated core (Core 0). On the other hand, the DNN model's L2
cache-miss rates did not increase as much. This suggests that
the DNN model's exeuction increase cannot be fully explained by
increased L2 cache-misses. Instead, as we hypothesized in the previous
experiment, the increased memory pressure from the co-scheduled memory
intensive benchmarks is the primary cause of the DNN model's execution
time increase. Therefore, we conclude that DeepPicar's DNN model is
more senstive to DRAM access latency than L2 cache space.

This observation suggests that shared cache partitioning
techniques~\cite{Gracioli2015,Kim2016} may not be effective isolation
solutions for DeepPicar as its AI workload is more sensitive to memory
performance. Instead, memory controller focused isolation solutions,
either hardware or software-based ones (e.g.,~\cite{Guo2017,Yun2013}),
may be more important. Although our observation is made on a single
hardware platform running on a single DNN workload, we suspect that
many AI workloads may exhibit similar characteristics.

@@

First, we find that DeepPicar's computing platform,
the Raspberry Pi 3 Model B, offers adquate computing capacity to
perform real-time control of the RC car at 20 Hz frequency (or
50ms per control loop). Given the complexity of the DNN we use, we
were pleasantly suprised by this finding. The time breakdown shows that
the DNN inferencing operation, performed by the Tensorflow library,
dominates the execution time, which is expected.

Second, we find that scalability of Tensorflow's DNN inference 
implementation is limited: We find that using all four cores is 
only about 30\% better than using just a single core.

>>>>

For real-time control of a car (or any robots), the control loop
frequency must be sufficiently high so that the car can quickly
react to the changing environment and its internal states. In general,
control performance improves when the frequency is higher, though
computation time and the type of particular physical system are
factors in determining a proper control loop latency. While a standard
control system may be comprised of multiple control loops with
differing control frequencies---e.g., a inner control loop for lower-level
PD control, a outer loop for motion planning, etc.---DeepPicar's
control loop is a single layer, as shown earlier in
Figure~\ref{fig:controlloop}, since a single deep neural network
replaces the traditional multi-layer control pipline. (Recall
Figure~\ref{fig:end-to-end-control} on the differences between the
standard robotics control vs. end-to-end deep learning approach).
This means that the DNN interference operation must be completed
within the inner-most control loop update frequency. To understand
achievable control-loop update frequencies, we experimentally measured
the execution times of DeepPicar's DNN inference operations.

@@

First, as expected, we find that the inference operation
dominates the control loop execution time, accounting for about 85\% of
the execution time.
Second, and more importantly, we also find that the measured average
execution time of a single control loop is 42.67 ms, or 23.4 Hz and
the 99 percentail time is 50.70 ms.
This means that the DeepPicar can operate
at about 20 Hz control frequency in real-time using only the on-board
Raspberry Pi 3 computing platform, without needing any remote computing
resources. We consider these results respectable given the complexity
of the deep neural network, and the fact that the inference operation
performed by Tensorflow only utilizes the CPU cores of the
Raspberry Pi 3 (its GPU is not supported by Tensorflow.)

@@

In this experiment, we investigate the scalability of performing
inference operations of DeepPicar's neural network with respect to the
number of cores. As noted earlier, the Raspberry Pi 3 platform has
four Cortex-A53 cores and Tensorflow 
provides a programmable mechanism to adjust how many cores are to be
used by the library. Leveraging this feature, we repeat the
same experiment described in the previous subsection with varying
numbers of CPU cores---from one to four.

@@

Figure~\ref{fig:perf-vs-corecnt} shows the average execution time of
the control loop as we vary the number of cores used by
Tensorflow. As expected, as we assign more cores, the average execution
time decreases---from 61.96 ms on a single core to 42.67 ms on four
cores (a 30\% improvement). However, the improvement is far from ideal
linear scaling. In particular, from 2 cores to 3 cores, the
improvement is mere 2.38 ms (or 4\%). In short, we find that the
scalability of DeepPicar's deep neural network is not ideal on the
platform. We do not know whether it is due to the limitations of
Tensorflow's multicore implementation or if it's the model's inherent
characteristics.

The poor scalability opens up the possibility of consolidating
multiple different tasks or different neural network models rather
than allocating all cores for a single neural network model. For
example, it is conceivable to use four cameras and four different
neural network models, each of which is trained separately and
executed on a single dedicated core. Assuming we use the same network
architecture for all models, then one might expect to achieve up to
15 Hz using one core (given 1 core can deliver 62 ms average
execution time). In the next experiment, we investigate the
feasibility of such a scenario.

@@

Figure~\ref{fig:perf-vs-modelcnt} shows the results. In the figure, the
X-axis shows the system configuration: \#of DNN models x \#of CPU
cores/DNN. For example, '4x1' means running four DNN models each of
which is assigned to run on one CPU core, whereas '2x2' means running
two DNN models, each of which is assigned to run on two CPU
cores. The Y-axis shows the average inference timing.
The two bars on the left show the impact of co-scheduling four DNN
models. Compared to executing a single DNN model on one CPU core
(1x1), when four DNN models are co-scheduled (4x1), each model
suffers an average inference time increase of approximately 15 ms,
$\sim$24\%. On the other hand, when two DNN models, each using two CPU
cores, are co-scheduled, the average inference timing is increased by
about 7 ms, or 10\%, compared to the baseline of running one model
using two CPU cores.

@@

These increases in inference times in the co-scheduled scenarios are
expected and likely caused by contention in the shared hardware
resources, such as the shared L2 cache and/or the DRAM controller.
To further analyze the source of contention, we use hardware
performance counters of the processor. Specifically, we measure L2
miss rates of the DNN models first in isolation and then after
co-scheduling other models. If the shared L2 cache is the primary
source of inteference, then the measured L2 miss rates will
increase. Figure~\ref{fig:l2missrate-vs-modelcnt} shows the results.
As can be see in the figure, L2 miss rates are largely unchanged
regardless of whether multiple models are co-scheduled or not. This
suggests that the shared L2 cache is not the bottleneck that caused
execution time increases. In other words, DNN models don't appear to be
sensitive to the shared L2 cache space. % L2 partitioning is probably
                                % not going to be useful.
Instead, we hypothesize that it is likely caused by the memory
controller---the only other major shared hardware source---where
memory requests from different CPU cores contend, which would result
in increased memory access latency. While some Intel processors
provide oncore hardware counters that can measure average memory
access latency~\cite{ye2016maracas}, we were not able to identify
whether such hardware counters exist in the BCM2837 processor of
Raspberry Pi 3 due to the lack of documentation. Instead, in the next
experiment, we use memory intensive synthetic benchmarks to test the
hypothesis.

@@

Figure~\ref{fig:} shows the normalized execution time and L2 miss-rate
of the DNN model running on the Core 0 as a function of the number of
co-scheduled memory intensive synthetic benchmarks. First, as we
increase the number of co-runners, the DNN model's execution times are
increased---by up to 9.6X---even though the DNN model is running on a
dedicated core (Core 0). On the other hand, the DNN model's L2
cache-miss rates do not increase as much. This suggests that
the DNN model's exeuction increase cannot be fully explained by
increased L2 cache-misses. Instead, as we hypothesized in the previous
experiment, the increased memory pressure from the co-scheduled memory
intensive benchmarks is the primary cause of the DNN model's execution
time increase. Therefore, we conclude that DeepPicar's DNN model is
more senstive to DRAM access latency than L2 cache space.

This observation suggests that shared cache partitioning
techniques~\cite{Gracioli2015,Kim2016} may not be effective isolation
solutions for DeepPicar, as its AI workload is more sensitive to memory
performance. Instead, memory controller focused isolation solutions,
either hardware or software-based ones (e.g.,~\cite{Guo2017,Yun2013}),
may be more important. Although our observation is made on a single
hardware platform running on a single DNN workload, we suspect that
many AI workloads may exhibit similar characteristics.

@@

First, we find that DeepPicar's computing platform,
the Raspberry Pi 3 Model B, offers adquate computing capacity to
perform real-time control of the RC car at 20 Hz frequency (or
50ms per control loop). Given the complexity of the DNN used, we
were pleasantly suprised by this finding. The time breakdown shows that
the DNN inferencing operation, performed by the Tensorflow library,
dominates the execution time, which is expected.

Second, we find that scalability of Tensorflow's DNN inference 
implementation is limited. We find that using all four cores is 
only about 30\% better than using just a single core.

### 2017-10-11 paper/intro.tex
Minor changes to paper

% related work and remaining problems
To understand what kind of computing hardware is needed for AI
workloads, we need a testbed and realistic workloads. While using a real car
based testbed would be most ideal, it is not only highly expensive, but also
poses serious safety concerns that hinder development and exploration.
Therefore, there is a strong need for safer and less costly
testbeds. There are already several relatively inexpensive RC-car
based testbeds, such as MIT 
RaceCar~\cite{shin2017project} and UPenn's F1$/$10 racecar~\cite{upennf1tenth}.
However, these RC-car testbeds still cost more than \$3,000, requiring
considerable investment.

>>>>

% related work and remaining problems
To understand what kind of computing hardware is needed for AI
workloads, we need a testbed and realistic workloads. While using a real 
car-based testbed would be most ideal, it is not only highly expensive, but also
poses serious safety concerns that hinder development and exploration.
Therefore, there is a strong need for safer and less costly
testbeds. There are already several relatively inexpensive RC-car
based testbeds, such as MIT 
RaceCar~\cite{shin2017project} and UPenn's F1$/$10 racecar~\cite{upennf1tenth}.
However, these RC-car testbeds still cost more than \$3,000, requiring
considerable investment.
